{\huge Supplements for: Changes and trends in budburst and leaf flush across Europe and North America} \\A meta-analysis of local adaptation in spring phenology studies

### 2024-02-21 Docs/manuscriptRev.Rnw
tracking changes -> diff file
A couple small tweaks throughout for flow, clarity

\begin{itemize}
  \item Forecasting the biological impacts of climate change requires understanding how species respond to warmer temperatures through inter-annual flexible variation vs. through adaptation to local conditions. Yet, we often lack this information entirely or find conflicting evidence across previous studies.
  \item We synthesize common garden studies across Europe and North America that reported spring event dates for a mix of angiosperm and gymnosperm tree species in the northern hemisphere, capturing data from 384 North American and 101 European provenances (i.e. populations) with observations from 1962 to 2019, alongside fall event data when provided.
  \item Across continents, we find no evidence of provenance effects (i.e. population differences) in spring phenology, but strong clines with latitude and mean annual temperature (MAT) in fall. These effects, however, appear to diverge by continent and species type (gymnosperm vs. angiosperm), with particularly pronounced clines in North America in fall.
  \item Our results suggest flexible, likely plastic responses, in spring phenology with warming, and potential limits---at least in the short term---due to provenance effects for fall phenology. They also highlight that, after over 250 years of common garden studies on tree phenology, we still lack a holistic predictive model of clines across species and phenological events.
\end{itemize}
% Forecasting the biological impacts of climate change requires understanding how species respond to warmer temperatures through inter-annual flexible variation versus through adaptation to local conditions. Yet, we often lack this information entirely or find conflicting evidence across studies. The latter is the case for shifts in spring phenology---one of the most reported and consistent impacts of anthropogenic climate change, and also one of the most critical to forecasting, given its role in carbon sequestration. Some common garden studies have found evidence of important provenance effects (i.e. population differences), which suggest there may be local adaptation in the underlying cues of spring phenology and mirrors findings for fall events, while other studies find no evidence. Here, we synthesize common garden studies across Europe and North America that reported spring event dates for a mix of angiosperm and gymnosperm tree species in the northern hemisphere, capturing data from 384 North American provenances and 101 European provenances with observations from 1962 to 2019, alongside fall event data when provided. Across continents, we find no evidence of provenance effects in spring phenology, but strong clines with latitude and mean annual temperature (MAT) for fall events. These effects, however, appear to diverge by continent and species type (gymnosperm versus angiosperm), especially for fall events where clines with latitude and MAT are much stronger in North America. Our results suggest flexible, likely plastic responses, in spring phenology with warming, and potential limits---at least in the short term---due to provenance effects for fall phenology. They also highlight that, after over 250 years of common garden studies on tree phenology, we still lack a holistic predictive model of clines across species and phenological events. 
% \end{abstract}

@@

Overall, spring events such as budburst and leaf flush were not related to provenance latitude or MAT, neither across continents (latitude: \Sexpr{format(round(mean(all_species_springpost_fitC_spring_lat),digits = 2), nsmall=2)} days/degree [\Sexpr{format(quantile(all_species_springpost_fitC_spring_lat, 0.05), digits = 1)} - \Sexpr{format(quantile(all_species_springpost_fitC_spring_lat, 0.95), digits = 2)}]; MAT: \Sexpr{format(round(mean(all_species_springpost_fitC_spring_mat),digits = 2), nsmall=2)} days/$^{\circ}$C [\Sexpr{format(quantile(all_species_springpost_fitC_spring_mat, 0.05), digits = 2)} - \Sexpr{format(quantile(all_species_springpost_fitC_spring_mat, 0.95), digits = 2)}]) (Fig.\ref{figure:springfall_latmat}, Table.S\ref{supplement-table:model_spring_lat} \& S\ref{supplement-table:model_spring_mat} in Supporting Information), nor within North America (latitude: \Sexpr{format(round(mean(northamerican_springpost_fitC_spring_lat),digits = 2), nsmall=2)} days/degree [\Sexpr{format(quantile(northamerican_springpost_fitC_spring_lat, 0.05), digits = 1)} - \Sexpr{format(quantile(northamerican_springpost_fitC_spring_lat, 0.95), digits = 2)}]; MAT: \Sexpr{format(mean(northamerican_springpost_fitC_spring_mat),digits = 1)} days/$^{\circ}$C [\Sexpr{format(quantile(northamerican_springpost_fitC_spring_mat, 0.05), digits = 2)} - \Sexpr{format(quantile(northamerican_springpost_fitC_spring_mat, 0.95), digits = 2)}]) or Europe (latitude: \Sexpr{format(round(mean(european_springpost_fitC_spring_lat),digits = 2),nsmall=2)} days/degree [\Sexpr{format(quantile(european_springpost_fitC_spring_lat, 0.05), digits = 2)} - \Sexpr{format(quantile(european_springpost_fitC_spring_lat, 0.95), digits = 2)}]; MAT: \Sexpr{format(round(mean(european_springpost_fitC_spring_mat),digits = 2),nsmall=2)} days/$^{\circ}$C [\Sexpr{format(quantile(european_springpost_fitC_spring_mat, 0.05), digits = 2)} - \Sexpr{format(quantile(european_springpost_fitC_spring_mat, 0.95), digits = 2)}]) (Fig.\ref{figure:continental_spptype_lat}A \& \ref{figure:continental_spptype_effect}A). Results were similar using other distance metrics in lieu of latitude (see Fig.S\ref{supplement-figure:lat_distance} for results using the difference between provenance and garden latitude, and the spherical distance between provenance and garden).
% Spring events advanced slightly with provenance latitude in Europe spring events are slightly earlier where provenance MAT is lower (i.e., higher, more northern latitudes), though uncertainty intervals strongly overlapped zero. --- no longer the case

@@

We found the coarse metrics of provenance latitude and MAT were generally good predictors of phenology, performing better than our more complex and data-rich metric of climate overlap. Latitude and MAT appeared to well represent how similar the climates are between the provenances and gardens in the temperate and boreal forest species we studied, yielding similar results to metrics calculated specifically in the spring with daily climate data. %EMWSep4 -- also avoid repeating how narrow the data is etc. -- it's good to point out limitations, but repeating them too much is not helpful -- this is the biggest dataset for this question on earth, we need to focus on what we learned from it, not so much on the limitations. Also, Europeans would not call all of the forests northern.

@@

%EMWSep30: Please double-check we're explaining trends correctly before you submit; I think we are -- but please check closely.
%EWMFeb5 -- more edits here about angio/gymno
Our results highlight that spring events show much weaker clines than fall events in tree species, but suggest important variation between species types in spring events. Angiosperms tended to budburst earlier in provenances that were warmer and more southerly, while gymnosperms trended in the opposite direction. Such differences could be driven by the varying investment strategies, given that all our studied gymnosperms were evergreen and all angiosperms were deciduous. As evergreen species photosynthesize without leafout they generally leafout much later than deciduous species, after most risk of spring frost \citep{panchen14}, and thus may avoid frost risk. In contrast, deciduous species may tend towards earlier leafout in warmer climates to compete best for access to light and other resources \citep{cat2019}. Testing these hypotheses would require more information on frost risk and forest community assembly from across the provenance locations, but seems an important step towards understanding the drivers of this variation. Without a clear mechanism, extrapolating these results to other species or across Europe and North America may be difficult, especially given biases in the data, and the distinct climatic, geographical, and ecological contexts of Europe and North America. Without a clear and universal mechanism driving the observed phenological responses, the direct applicability of our findings to other contexts must be approached with caution. %EWMFeb5 -- trying to address request by R2 about extrapolating while also keeping ms flow. 
%EMWSep4 -- I like the below (and I think it may have been my idea), but we'd likely be asked to test for it ... and I am not 100% sure it makes sense -- both angiosperms and conifers need to activate their xylem to photosynthesize I believe. 
% Meanwhile, xylem anatomy also plays a part in determining spring timing due to the importance of vessel diameter to embolism risks induced by frost in spring. According to \citet{salk20}, species with narrow xylem such as coniferous gymnosperms and diffuse-porous angiosperms did not exhibit the consistent phenological trends shown in ring-porous species. Therefore, there might be additional differences if we have more species and more information on leafout timing and frost risk approach.

@@

\clearpage
\section{Figures}
%alinaSep22: i dont know how to force all figures to be in this section instead of in between texts
%EMWSep30 -- you just move them here! Please check I put them in the right order; I also added a \clearpage which can help with formatting looking clean (see if it works, you can remove if not). 
\begin{figure}[!h] 
    \centering
 \includegraphics[width=\textwidth]{..//..//localadaptclim/Docs/figure_ms/map_gardens.jpg}
    \caption{Distribution of common gardens (denoted as stars) and provenances (denoted as circles) included in this meta-analysis. The distribution was skewed toward North America (12 North American studies versus 5 European studies). See Table.S\ref{supplement-table:all_studies} in Supporting Information for sourcing information on selected studies. Note: map lines do not necessarily depict accepted national boundaries. European sites are made bold and denoted by an asterisk (*). }
    \label{figure:map_gardens}
\end{figure}

\begin{figure}[!h] 
    \centering
 \includegraphics[width=\textwidth]{..//..//localadaptclim/Docs/figure_ms/springfall_latmat.jpg}
    \caption{Event day of year (DOY) in relation to provenance latitude and MAT, coded by symbol for species and color for garden with linear fits from hierarchical Bayesian models. Spring events shown on top and fall events at the bottom. European gardens and species are made bold and denoted by an asterisk (*). } 
    \label{figure:springfall_latmat}
\end{figure}

>>>>

\begin{itemize} %EMWFeb20 -- small tweaks to improve flow and we do not need to say 'population' so much, methinks. 
  \item Forecasting the biological impacts of climate change requires understanding how species respond to warmer temperatures through inter-annual flexible variation vs. through adaptation to local conditions. Yet, we often lack this information entirely or find conflicting evidence across studies, which is the case for spring phenology.
  \item We synthesize common garden studies across Europe and North America that reported spring event dates for a mix of angiosperm and gymnosperm tree species in the northern hemisphere, capturing data from 384 North American and 101 European provenances (i.e. populations) with observations from 1962 to 2019, alongside fall event data when provided.
  \item Across continents, we find no evidence of provenance effects in spring phenology, but strong clines with latitude and mean annual temperature (MAT) in fall. These effects, however, appear to diverge by continent and species type (gymnosperm vs. angiosperm), with particularly pronounced clines in North America in fall events.
  %EWMFeb5 -- if you need to save words, can cut 'due to provenance effects' below. 
  \item Our results suggest flexible, likely plastic responses, in spring phenology with warming, and potential limits---at least in the short term---due to provenance effects for fall phenology. They also highlight that, after over 250 years of common garden studies on tree phenology, we still lack a holistic predictive model of clines across species and phenological events.
\end{itemize}
% Forecasting the biological impacts of climate change requires understanding how species respond to warmer temperatures through inter-annual flexible variation versus through adaptation to local conditions. Yet, we often lack this information entirely or find conflicting evidence across studies. The latter is the case for shifts in spring phenology---one of the most reported and consistent impacts of anthropogenic climate change, and also one of the most critical to forecasting, given its role in carbon sequestration. Some common garden studies have found evidence of important provenance effects (i.e. population differences), which suggest there may be local adaptation in the underlying cues of spring phenology and mirrors findings for fall events, while other studies find no evidence. Here, we synthesize common garden studies across Europe and North America that reported spring event dates for a mix of angiosperm and gymnosperm tree species in the northern hemisphere, capturing data from 384 North American provenances and 101 European provenances with observations from 1962 to 2019, alongside fall event data when provided. Across continents, we find no evidence of provenance effects in spring phenology, but strong clines with latitude and mean annual temperature (MAT) for fall events. These effects, however, appear to diverge by continent and species type (gymnosperm versus angiosperm), especially for fall events where clines with latitude and MAT are much stronger in North America. Our results suggest flexible, likely plastic responses, in spring phenology with warming, and potential limits---at least in the short term---due to provenance effects for fall phenology. They also highlight that, after over 250 years of common garden studies on tree phenology, we still lack a holistic predictive model of clines across species and phenological events. 
% \end{abstract}

@@

Overall, spring events such as budburst and leaf flush, were not related to provenance latitude or MAT, neither across continents (latitude: \Sexpr{format(round(mean(all_species_springpost_fitC_spring_lat),digits = 2), nsmall=2)} days/degree [\Sexpr{format(quantile(all_species_springpost_fitC_spring_lat, 0.05), digits = 1)} - \Sexpr{format(quantile(all_species_springpost_fitC_spring_lat, 0.95), digits = 2)}]; MAT: \Sexpr{format(round(mean(all_species_springpost_fitC_spring_mat),digits = 2), nsmall=2)} days/$^{\circ}$C [\Sexpr{format(quantile(all_species_springpost_fitC_spring_mat, 0.05), digits = 2)} - \Sexpr{format(quantile(all_species_springpost_fitC_spring_mat, 0.95), digits = 2)}]) (Fig.\ref{figure:springfall_latmat}, Table.S\ref{supplement-table:model_spring_lat} \& S\ref{supplement-table:model_spring_mat} in Supporting Information), nor within North America (latitude: \Sexpr{format(round(mean(northamerican_springpost_fitC_spring_lat),digits = 2), nsmall=2)} days/degree [\Sexpr{format(quantile(northamerican_springpost_fitC_spring_lat, 0.05), digits = 1)} - \Sexpr{format(quantile(northamerican_springpost_fitC_spring_lat, 0.95), digits = 2)}]; MAT: \Sexpr{format(mean(northamerican_springpost_fitC_spring_mat),digits = 1)} days/$^{\circ}$C [\Sexpr{format(quantile(northamerican_springpost_fitC_spring_mat, 0.05), digits = 2)} - \Sexpr{format(quantile(northamerican_springpost_fitC_spring_mat, 0.95), digits = 2)}]) or Europe (latitude: \Sexpr{format(round(mean(european_springpost_fitC_spring_lat),digits = 2),nsmall=2)} days/degree [\Sexpr{format(quantile(european_springpost_fitC_spring_lat, 0.05), digits = 2)} - \Sexpr{format(quantile(european_springpost_fitC_spring_lat, 0.95), digits = 2)}]; MAT: \Sexpr{format(round(mean(european_springpost_fitC_spring_mat),digits = 2),nsmall=2)} days/$^{\circ}$C [\Sexpr{format(quantile(european_springpost_fitC_spring_mat, 0.05), digits = 2)} - \Sexpr{format(quantile(european_springpost_fitC_spring_mat, 0.95), digits = 2)}]) (Fig.\ref{figure:continental_spptype_lat}A \& \ref{figure:continental_spptype_effect}A). Results were similar using other distance metrics in lieu of latitude (see Fig.S\ref{supplement-figure:lat_distance} for results using the difference between provenance and garden latitude, and the spherical distance between provenance and garden).
% Spring events advanced slightly with provenance latitude in Europe spring events are slightly earlier where provenance MAT is lower (i.e., higher, more northern latitudes), though uncertainty intervals strongly overlapped zero. --- no longer the case

@@

%EWMFeb20 -- tweak below, could cite these lines also for the response to reviewer about MAT
We found the coarse metrics of provenance latitude and MAT were generally good predictors of phenology, performing better than our more complex, data-rich, and season-specific metric of climate overlap. Latitude and MAT appeared to well represent how similar the climates are between the provenances and gardens in the temperate and boreal forest species we studied, yielding similar results to metrics calculated specifically in the spring with daily climate data. %EMWSep4 -- also avoid repeating how narrow the data is etc. -- it's good to point out limitations, but repeating them too much is not helpful -- this is the biggest dataset for this question on earth, we need to focus on what we learned from it, not so much on the limitations. Also, Europeans would not call all of the forests northern.

@@

%EMWSep30: Please double-check we're explaining trends correctly before you submit; I think we are -- but please check closely.
%EWMFeb5 -- more edits here about angio/gymno
Our results highlight that spring events show much weaker clines than fall events in tree species, but suggest important variation between species types in spring events. Angiosperms tended to budburst earlier in provenances that were warmer and more southerly, while gymnosperms trended in the opposite direction. Such differences could be driven by the varying investment strategies, given that all our studied gymnosperms were evergreen and all angiosperms were deciduous. As evergreen species photosynthesize without leafout they generally leafout much later than deciduous species, after most risk of spring frost \citep{panchen14}, and thus may avoid frost risk. In contrast, deciduous species may tend towards earlier leafout in warmer climates to compete best for access to light and other resources \citep{cat2019}. Testing these hypotheses would require more information on frost risk and forest community assembly from across the provenance locations, but seems an important step towards understanding the drivers of this variation. Without a clear mechanism, extrapolating these results to other species or across Europe and North America may be difficult, especially given biases in the data, and the distinct climatic, geographical, and ecological contexts of Europe and North America.  
%EWMFeb20 -- deleted last sentence here, update letter
%EWMFeb5 -- trying to address request by R2 about extrapolating while also keeping ms flow. 
%EMWSep4 -- I like the below (and I think it may have been my idea), but we'd likely be asked to test for it ... and I am not 100% sure it makes sense -- both angiosperms and conifers need to activate their xylem to photosynthesize I believe. 
% Meanwhile, xylem anatomy also plays a part in determining spring timing due to the importance of vessel diameter to embolism risks induced by frost in spring. According to \citet{salk20}, species with narrow xylem such as coniferous gymnosperms and diffuse-porous angiosperms did not exhibit the consistent phenological trends shown in ring-porous species. Therefore, there might be additional differences if we have more species and more information on leafout timing and frost risk approach.

@@

\clearpage
\section{Figures}
%alinaSep22: i dont know how to force all figures to be in this section instead of in between texts
%EMWSep30 -- you just move them here! Please check I put them in the right order; I also added a \clearpage which can help with formatting looking clean (see if it works, you can remove if not). 
\begin{figure}[!h] 
    \centering
 \includegraphics[width=\textwidth]{..//..//localadaptclim/Docs/figure_ms/map_gardens.jpg}
    \caption{Distribution of common gardens (denoted as stars) and provenances (denoted as circles) included in this meta-analysis. The distribution was skewed toward North America (12 North American studies versus 5 European studies). See Table.S\ref{supplement-table:all_studies} in Supporting Information for sourcing information on selected studies. Note: map lines do not necessarily depict accepted national boundaries. European sites are shown in bold and denoted by an asterisk (*). }
    \label{figure:map_gardens}
\end{figure}

\begin{figure}[!h] 
    \centering
 \includegraphics[width=\textwidth]{..//..//localadaptclim/Docs/figure_ms/springfall_latmat.jpg}
    \caption{Event day of year (DOY) in relation to provenance latitude and MAT, coded by symbol for species and color for garden with linear fits from hierarchical Bayesian models. Spring events shown on top and fall events at the bottom. European gardens and species are shown in bold and denoted by an asterisk (*). } 
    \label{figure:springfall_latmat}
\end{figure}
